Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2020-10-24 manu_draft.Rmd
workingin PErry methods

### General approach

The methods proposed here involve three stages.  In the first stage, a GAM is used to estimate a smooth pattern of variation in the raw data along with its uncertainty.  In the second stage, a feature of interest is calculated from the estimated GAM, along with its propagated uncertainty.  In the examples here, the feature is a seasonal average.  Other features could be the timing or magnitude of a seasonal peak, but those are not developed here.  In the third stage, a mixed effects meta-analysis is used to estimate trends and/or test hypotheses about the change in seasonal averages across years.  While meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty [@Gasparrini12;@Sera19].  They use response data that come with standard errors (uncertainties), which is exactly the need here.  We give two kinds of mixed effects meta-analysis steps: simple comparison of whether seasonal features differ across years, and estimation of short-term linear trends on time-scales chosen by the analyst.  These allow analysis of new kinds of questions for long-term water quality monitoring data.

The three-stage approach is motivated in several ways.  In the first stage, the GAM serves the role of signal-extraction at the scale of the raw data.  We explain the relationships among various GAM specifications and how some previous methods have entangled the trend-analysis goal with the signal-extraction goal in the formulation of splines, making it harder to interpret either one.  The second stage harnesses the capability to propagate uncertainty from estimated GAM parameters to features of interest such as seasonal averages.  Such features are usefully viewed as functions of the fitted GAM rather than parameters to be estimated within the GAM.   On a practical level, the first two stages can be viewed as creating a "data product" that can be used in any subsequent analyses as long as they incorporate the uncertainty associated with each point estimate. This offers an important practical benefit for visualization and management purposes. For the third stage (long-term trend-analysis), we harness existing meta-analysis methods that are well-suited to the task at hand.

The basic structural form of a Generalized Additive Model is [@Hastie90]:

\begin{equation}
E\left(Y|X_1, X_2, ..., X_p\right) = s_0 + \sum_{j=1}^{p} s_j\left(X_j\right) 
(\#eq:gengam)
\end{equation}

where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an intercept term $s_0$.  The smoothing functions are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.  Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.  All methods herein use functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods [@Wood17].

To evaluate GAMs with SFE chlorophyll data, we followed an approach that built on past work by the Chesapeake Bay Program [@Murphy19].  We specifically build our models using the `gam1` and `gam2` structures in @Murphy19, but present a different naming convention to better distinguish the functional components of the models.  The `gam1` and `gam2` models are shown in equations \@ref(eq:gamone) and \@ref(eq:gamtwo), denoted as "Constant season" and "Interactive season", respectively.  We also describe the "Constant season*" model below as a slight modification of the former. Each model has a similar functional form for time in relation to chlorophyll with slight differences in the smoothers that describe time.

\begin{equation}
\boldsymbol{Constant\ season:} g\left(chl\right) = s_0 + \beta_1 decYear + s_1\left(doy\right) + s_2\left(decYear\right) + \epsilon 
(\#eq:gamone)
\end{equation}

\begin{equation}
\boldsymbol{Interactive\ season:} g\left(chl\right) = s_0 + \beta_1 decYear + s_1\left(doy\right) + s_2\left(decYear\right) + ti_1(doy, decYear) + \epsilon
(\#eq:gamtwo)
\end{equation}

For each GAM, chlorophyll was modelled using a link function $g()$ that relates the response variable to the independent variables $decYear$ and day of year ($doy$) (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).  The $decYear$ variable is a continuous measure for date in decimal time, where date is expressed as a continuous numeric value with year on the left of the decimal and date within the year to the right of the decimal (e.g., July 1st 2019 would be 2019.5).  The $doy$ variable is an integer value for the Julian day in each year (from 1 to 366) that represents the seasonal component of the time series. The estimate $s_0$ is an intercept value and $\beta_1$ is a slope estimate for a linear year effect included in all models. The $s_1()$ and $s_2()$ functions are different smoothers for $decYear$ and $doy$ where the former uses a thin-plate spline and the latter uses a cyclic cubic spline to capture the periodic seasonal component [@Wood03].  The $ti_1()$ function is a tensor product interaction between $decYear$ and $doy$ that allows the seasonal component to change between years.  Within the $ti_1()$ function are the two separate smoothers for year as a thin-plate spline and $doy$ as a cyclic cubic spline. As with conventional linear models, the remaining variance in the response not captured by the functional form of the model is expressed as an error term $\epsilon$ that is normally-distributed having mean equal to zero and variance as $\sigma^2$.  The residual variance is used to back-transform model estimates for trend testing, described below.

A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., equation \@ref(eq:gamone) or \@ref(eq:gamtwo)) and determining how much variation could be explained by the smoothers in each function.  The two models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, and $ti_1()$) require user input on the potential upper limit for how much variation could be explained by the predictors during model fitting.  Each smoother is approximated by multiple, localized spline functions that are connected by knots (*k*-values) to create a continuous function. Increasing the number of knots creates a smoother fit of the response variable against the predictor, whereas choosing a small number of knots creates a more rigid function that may underfit the data.  Initial testing by @Murphy19 showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in chlorophyll data from Chesapeake Bay.  The upper limit for the number of knots for the $s_2(decYear)$ smoother was modified from the default value to the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.  We follow the same approach herein both models.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, an extension of the constant season model was also added as "Constant season*", where the total number of knots for the $decYear$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $decYear$ to be unconstrained in how much variance is fit within and between years.

The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.  A comparison of GAM performance between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.  The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable. However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisfying assumptions for GAMs.  The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation [@Box64].  The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.  The profile vector for each chlorophyll time series was obtained using the *boxcox* function from the MASS package [@Venables02] to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.  Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation [@Box64]:

\begin{equation}
g\left(chl;\lambda\right) = \begin{cases} \frac{chl ^ {\lambda} - 1}{\lambda} & \lambda \ne 0 \\ \log\left(chl\right) & \lambda = 0 \end{cases}
(\#eq:boxcox)
\end{equation}

where $g()$ is the transformation function for the appropriate GAM at the estimated $\lambda$ value. Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.  The *BoxCox* function from the forecast package [@Hyndman20] in R was used to transform the chlorophyll time series with the optimal $\lambda$ value.  Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.

All models were compared using standard summary statistics describing overall model fit to the observed data.  In addition to the Generalized Cross-Validation score provided by the mgcv package, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.  Comparisons were made between the different GAMs and for the transformations used to define the $g()$ function for chlorophyll at each station. Model predictions were based on the standard model output from the *predict()* function from the mgcv package.  A long-term trend independent of seasonal variation was also estimated for each model by subtracting the seasonal terms (i.e., $s_1(doy)$ and $ti_1(doy, year)$) from the predictions. Back-transformation of results were obtained by exponentiation if the log-transformed variable was used or using the *InvBoxCox()* function from the forecast package if the Box-Cox transformation was used. For hypothesis tests that required estimates of central tendency and confidence intervals for mean values, back-transformations were based on an exponentiation or inverse function of the transformed response variable as before, with the addition of a correction factor.  This bias-correction factor was simply the variance of the model residuals divided by two for variables in log-space and correction with a forecast variance estimate as specified in the *InvBoxCox()* function from the forecast package.

## Using GAMs to evaluate trends

Each of the GAM structures provided different approaches to quantify long-term variation in annual and seasonal changes.  Although useful as a general approach to view functional forms for different time series, additional information on whether an observed change is statistically significant can be useful to inform environmental management decisions.  We present two methods for performing trend tests on model results that address different changes that may be of management interest.  First, a method for evaluating the percent change in annual averages between two time periods is briefly summarized [@Murphy19].  Second, a method for evaluating means and confidence intervals from GAM results for seasonal periods is presented, including an estimate of direction and rate of seasonal change over time.

First, for two time periods of interest, trends were estimated as the difference in means from the GAM results between the periods, an estimate of percent change, and an estimate of statistical certainty of the comparison.  @Murphy19 defined the two time periods as "base" and "test" periods, where the former represents a baseline condition earlier in the time series as a basis of comparison to the more recent "test" period.  The differences are estimated as:

\begin{equation}
\text{Difference} = Z_d\hat{\beta}
(\#eq:diffest)
\end{equation}

where $Z_d$ is a point estimate of the differences of the predictions between the base and test years and $\hat{\beta}$ is the estimated parameter vector from the fitted GAM.  The standard error of this difference is then:

\begin{equation}
\text{se}_{\text{diff}} = \sqrt{Z_d\hat{\Sigma}_\beta Z_d^T}
(\#eq:sediffeest)
\end{equation}

defined as the square root of the product of $Z_d$, the estimated variance-covariance matrix of parameter vector for the fitted GAM $\hat{\Sigma}_\beta$, and the transposition of $Z_d$.  The trend is considered significant if the appropriate confidence interval at a specified $\alpha$ does not include zero.

The second trend test evaluated changes in seasonal estimates between years using results from a fitted GAM.  This test provided both an estimate of a seasonal mean, its associated confidence interval, and if the rate of change in the seasonal means has varied over a specific period of interest.  For example, changes that occur during late fall or early spring can be evaluated between years to determine if the seasonal averages differ and if the direction and magnitude of change is significant. This has implications for understanding potential changes in seasonal bloom phenology resulting from shifting system responses to nutrient inputs or changes in the latter.

Seasonal averages and confidence intervals were estimated from fitted GAMs by first creating a matrix of the predictors (i.e., `decYear`, `doy`) covering defined seasons in the time series.  For example, seasonal changes from late summer could be obtained from a prediction matrix covering all dates from Aug 1st to Sep. 30th for the period from 2000 to 2010.  The prediction matrix is then used to obtain fitted values from a GAM that describe the estimated values of chlorophyll that occurred during the periods of interest. Rather than the estimated values for the predictions, a linear predictor matrix was obtained that provided a parameter vector for the supplied covariate values.  This predictor matrix was then used to estimate the mean seasonal value within a year and associated variance using the sum of predictions from the matrix.  As a result, each year had a mean value of chlorophyll with an associated confidence interval that describes the modelled expectation for the seasonal period.

Although the seasonal averages obtained from GAMs described the expected values for each year, describing the direction and magnitude of the seasonal change across years was also of interest.  Conventionally, trend analysis of changes in a response variable could be achieved with ordinary least squares regression, where the response variable is the expected seasonal value of chlorophyll and the predictor is year as a numeric value.  Although this approach could be used to estimate a seasonal rate of change, the linear fit through the seasonal means does not account for the variance in the seasonal averages.  To address this issue, we use mixed-effects regression meta-analysis (mixed-meta hereafter) to evaluate seasonal changes and their uncertainty over a finite set of years.  This analysis is an extension of conventional regression by accounting for uncertainty attributed to both variation of the true seasonal mean around the trend and variation of the estimated seasonal mean around the true seasonal mean:

\begin{equation}
chl_i = \beta_0 + \beta_1 year_i + b_i + \epsilon_i
(\#eq:mixmet)
\end{equation}

where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and deviation of the seasonal mean from the true mean ($S_i$).  Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and variance for $\epsilon_i$ equal to the squared standard error from the seasonal chlorophyll averages.  The R package *mixmeta* was used to estimate trends in the seasonal averages over time [@Sera19].  Results have a similar interpretation as those from regression analysis (i.e., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.

## Model application and interpretation

Each of the three GAM structures (constant season, interactive season, and constant season*) were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.  For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoother terms.  Models were also evaluated using different transformations to determine which had improved fit for specific time series.  Trend analyses with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.  Temporal and spatial patterns within and between stations were also identified to better quantity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.

@@

>>>>

The methods proposed here involve three stages.  First, a GAM is used to estimate a smooth pattern of variation in the raw data along with its uncertainty.  Second, a feature of interest is calculated from the estimated GAM, along with its propagated uncertainty. For this example, the seasonal averages are extracted, whereas other features could be the timing or magnitude of a seasonal peak, but those are not developed here.  Third, a mixed effects meta-analysis is used to estimate trends and test hypotheses about the change in seasonal averages across years.  While meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty [@Gasparrini12;@Sera19].  Meta-analysis uses response data that includes standard errors (uncertainties) as needed to address our questions.  Two mixed effects meta-analysis approaches are developed, 1) simple comparison of whether seasonal features differ across years, and 2) estimation of short-term linear trends on time-scales chosen by the analyst.

The three-stage approach is motivated in several ways.  In the first stage, the GAM is used for signal-extraction at the scale of the raw data and we explain the relationships among various GAM specifications to emphasis concordance between the trend-analysis goal and signal-extraction goal in the formulation of splines.  The second stage uses mixed-meta analysis to propagate uncertainty from estimated GAM parameters to features of interest, i.e., seasonal averages.  Such features can be practically evaluated as functions of the fitted GAM, rather than as parameters to be estimated within the GAM. On a practical level, the first two stages can be viewed as creating a "data product" that can be used in any subsequent analyses as long as they include the uncertainty associated with each point estimate. This offers an important practical benefit for visualization and management purposes. For the third stage (long-term trend-analysis), we harness existing meta-analysis methods that are well-suited to the task at hand.

### First-stage analysis: GAM estimation

Generalized additive models use sums of parametric terms (e.g., linear) and non-parametric terms (e.g., smoothing splines) to predict response data. In our approach, GAMs are used to estimate signal and uncertainty from the time-series data.  Results are subsequently used in the second-stage analysis, described below.

In this section we describe four GAM formulations that can achieve the same or similar fits to data in slightly different ways.  We present these to clarify relationships that may otherwise be confusing and to emphasize that results can be similarly achieved if a model unconstrained to have sufficient capability to extract a signal from raw data. Models will be written in the notation of the `mgcv` R package as formulas for the `gam` function [@Wood17].

The fundamental goal is to smooth the raw data across time in order to separate the signal from the noise. The simplest GAM for this purpose can be expressed as:

Model S: `y ~ s(cont_year, k = num_knots_Y)`

Here `y` is the time-series of interest, such as Chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knots used to create the spline.

For smoothing in `mgcv`, appropriate smoothness is achieved not by choice of knots but by a penalty on the net curvature of the spline [@Wood04].  Smoothing was determined using generalized cross-validation (GCV, as implemented in `mgcv`), which approximately minimizes out-of-sample prediction error.  To allow GCV (or other alternatives) to work as intended, the number of knots must be sufficiently large.  Results should not be sensitive to the number of knots; if they are, the number of knots should be increased.  Therefore, we choose a sufficiently large number of knots for `num_knots_Y` as 12 times the number of years in the time series that was modeled.  This created the potential to have one knot per month as an approach to both prevent under-fitting the observed data and to accurately estimate the seasonal signal within a year.  In some cases, this upper limit was too large given the observed data and the multiplier was reduced by one until sufficient degrees of freedom were available (i.e., 12 * years, 11 * years, etc.).

The next three spline formulations differ in how spline terms compose a model to smooth the raw data.  For convenience, a linear trend in `cont_year` can specified explicitly as a separate component.  This would be expressed as:

Model SY: `y ~ cont_year + s(cont_year, k = num_knots_Y)`

This model is effectively mathematically equivalent to model S.  The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.  When `cont_year` is included explicitly as a linear term in model SY, the `mgcv` implementation automatically adjusts the basis functions for the spline to exclude the linear term, to avoid over-parameterizing the model.  With model S, the estimated trend in `cont_year` and its uncertainty can be extracted from the fitted spline.  With model SY, it is available more conveniently.  Further, package `mgcv` includes an option `select = TRUE` to penalize linear trends in splines to provide a method for variable selection, such as when numerous splines are included in the model formulation for variables that may or may not be important.  For our approach, this option is not used and all models specify `select = FALSE`.  Details in the supplement explain this justification.

Next, an average within-year cyclic pattern as a separate spline can be separted in the model formulation.  This can be expressed as:

Model SYD: `y ~ cont_year + s(cont_year, k = num_knots_Y) + s(doy, bs = 'cc', k = num_knots_D)`

Here `doy` is "day-of-year" (i.e., Julian date, a count starting January 1 for each year), `bs = 'cc'` indicates the spline will be cyclic (constrained to start and end at the same value), and `num_knots_D` is the upper limit for the number of knots for the `doy` spline.  While model SYD is not mathematically equivalent to models S and SY, it should be functionally similar.  In this model, the `doy` spline gives the average pattern of fluctuations within each year.  This changes the interpretation of the `cont_year` spline to represent smoothed deviations from the average within-year pattern.

Models S, SY, and SYD can all potentially extract a similar signal from the raw data.  What differs is the allocation of penalties for curvature used to determine smoothness for each spline.  In model SYD, there are separate penalties for the two splines.  Caution is needed in interpreting differences in fit between model SYD and models S or SY because the penalties for smoothing splines based on curvature are heuristic [@Wood17]. For example, if a lower AIC is achieved in one model compared to another, assuming both use sufficient knots, this may just reflect the outcome of alternative penalization heuristics implied by the different formulations. In the examples here, model SYD achieves nearly identical fits to model S or SY, where the latter by definition also achieve identical fits.

Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will "borrow information" across years.  The model will impute an average seasonal pattern where data are limited. However, a conceptual representation of what an average within-year pattern represents is challenging.  For example, suppose in every year includes a spring chl-a peak.  If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude and appeas as a "typical" year.  However, if the magnitude is the same and the timing varies across years, then the average within-year pattern will not appear typical.  The imputed values will represent a day-by-day average, appearing more spread out in time than the pattern in any single year.

More importantly, the average within-year pattern (and its uncertainty) can be obtained from estimated models S and SY, as well as from model SYD, but using a different approach.  In models S and SY, the average within-year pattern is simply the average across years of the smoothed within-year patterns.  From this perspective, model SYD, like model SY, is a convenience, estimating the average within-year pattern directly as a separate estimated spline.  The average pattern extracted from model S or SY may not be identical to the `doy` spline estimated in model SYD, again due to different smoothing formulations, but in practice they should be similar.

Finally, the raw data can be smoothed using a bivariate spline representing an interaction between `cont_year` and `doy`. This can be expressed as:

Model SYDI: `y ~ cont_year + s(cont_year) + s(doy, bs = "cc") + ti(cont_year, doy, bs = c("tp", "cc"), k = c(num_knots_Y_ti, num_knits_D_ti))`

Here, `ti()` specifies a tensor-product spline for a surface that varies smoothly as a function of both `cont_year` and `doy`.  The number of knots is the product of `num_knots_Y_ti` in the `cont_year` axis and `num_knots_D_ti` in the `doy` axis.  In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knits_D_ti` or a sufficiently large value for knots in `s(cont_year)`, but not both to avoid over-parameterization.

Following similar rationale above, the relationship of model SYDI to model S is similar to that of model SYD to model S.  Model SYDI is more mathematically different, but all of the splines use the same inputs to smooth the same data.  The splines in `dec_time` and `doy` will likely not capture as much variation as above because of fewer knots.  The `ti` term represents an interaction by allowing the pattern in `dec_time` to vary by `doy` and vice-versa.  The interaction term in model SYDI provides an appearance that this model is fundamentally different form those provided by the other models.  However, models S, SY, and SYD all allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time-series.  Although model SYDI is the only model that includes an explicit interaction term, all of the models support the interaction conceptually. By providing this term with sufficient knots, the raw data can be fully smoothed with model SYDI to a similar degree as for the other models.  However, this outcome is difficult to achieve for reasons explained below.

The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly across years.  For example, if one year has an early and large magnitude spring bloom, the years before and after would have similar patterns described by the model, perhaps different but not substantially. For the raw San Francisco Bay data, this may not be the case, as one year might have an early, high spring bloom and the next year a late, low spring bloom, or other such large variation across years.  If the interaction spline has sufficient knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.  However, if that is the case, the conceptual motivation for model SYDI may not be justified.  In summary, model SDYI may best be suited for systems where within-year fluctuations actually vary smoothly across years, in which case it may out-perform the other models.  If that is not the case, it may perform similarly to the others, but the spline structure and penalty structure are more different, making performance harder to compare.

All of the models require a sufficiently large number of knots in one or more of the spline terms to accommodate the fluctuations in the data and use the GAM estimation methods in the manner for which they were designed.  In models S, SY, and SYD, the number of knots in the `cont_year` term need to be sufficiently large.  In model SYDI, a very large number of knots in both the `cont_year` spline *and* in both dimensions of the interaction spline is impossible to achieve.  Allowing the `cont_year` spline to include many knots would somewhat defeat the purpose of the interaction spline.  In practice, allocation of knots between the different splines can be chosen by the analyst so long it is recognized that different choices will arbitrarily lead to more variation being explained between the splines.

@Murphy19 used a related set of spline formulations but used them differently than we do here. @Murphy19 have a "`gam0`" with only a `s(doy)` term, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.  Their `gam0` is *a priori* not of interest for our data and was not explored further.  Compared to our uses of SYD and SYDI, their uses differ in choice of knots and interpretation of results.  For SYD, they set a maximum number of knots in the `s(cont_year)` term of 2/3 times the number of years, whereas we use 12 times the number of years, roughly one per month.  Thus, they interpret this spline as separating a long-term (or low-frequency) trend from other patterns, whereas we use it to separate signal from noise at a scale informed by the data.  In the SYDI model, @Murphy19 do not explicitly consider number of knots in the interaction spline.  In both cases, @Murphy19 in effect use choice of number of knots in different spline components to serve as ad hoc allocation of variation in the data to different components while not necessarily allowing any component to fully estimate the signal in the data. @Murphy19 acknowledge that incomplete modeling of fluctuations in the data may lead to inflated Type I error rates, which would apply to their later comparisons of changes across time, but they leave that problem for future work.  For the methods presented here, we seek to avoid inflated Type I error rates arising in this way.  Finally, @Murphy19 present large AIC differences between their spline formulations.  We instead emphasize that, given sufficient knots, the models represent alternative formulations of conceptually similar explanations for the data and yield similar fits.  In our example, such large differences in AIC would only reflect inadequate choice of knots in one or more splines.

### Log and Box-Cox transformations

Following previous authors, we consider base 10 log-transformed chl-a levels for analysis.  To generalize this choice, we also consider Box-Cox transformations, which scale from log-transformation to no (i.e., identity) transformation as a parameter $\lambda$ scales from 0 to 1.

### Second-stage analysis: Uncertainty propagation from estimated GAMs to seasonal features

\newcommand{\hm}{\hat{\mu}}
\newcommand{\hmt}{\hat{\mu}_t}
\newcommand{\mt}{\mu_t}
\newcommand{\hmr}{\hat{\mu}_r}
\newcommand{\hshmt}{\hat{\sigma}_{\hat{\mu}, t}}
\newcommand{\hshmr}{\hat{\sigma}_{\hat{\mu}, r}}
\newcommand{\hshm}{\hat{\sigma}_{\hat{\mu}}}
\newcommand{\hsshm}{\hat{\sigma}^2_{\hat{\mu}}}
\newcommand{\hsshmt}{\hat{\sigma}^2_{\hat{\mu}, t}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\hby}{\hat{\mathbf{y}}}
\newcommand{\hbb}{\hat{\boldsymbol{\beta}}}
\newcommand{\hShb}{\hat{\Sigma}_{\hat{\beta}}}
\newcommand{\hShy}{\hat{\Sigma}_{\hat{\mathbf{y}}}}
\newcommand{\sbs}{\sigma_b^2}

Next we consider the goal of estimating a seasonal average, such as the mean spring Chl-A level, and its uncertainty in each year.  Define $\mu_t$ as the seasonal average in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$. The season includes $n$ days.  In what follows, subscript $t$ will be omitted.

Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of estimate parameters and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.  Here $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.  For example, using model SY, if a point estimate for Chl-A is needed for a single day, given as `dec_year` = $r$, $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns. The number of spline basis functions is related to the number of knots. Note that $r$ can be any time; it need not be the time of an observation.

To obtain a vector, $\hby$, of fitted point estimates for every day in a season, $\bX$ would have one row for each day.  The seasonal averages used in our examples are calculated at the resolution of days.  The estimated spline yields both $\hbb$ and $\hShb$, an estimate of the covariance matrix of the sampling distribution of $\hbb$.  The scalar standard errors of $\hbb$ are the square roots of the diagonal elements of $\hShb$.  Since parameter estimates are correlated, it is important to use the full $\hShb$ to determine the covariance of $\hby$.  This is given by $\hShy = \bX \hShb \bX^T$.

Finally, the estimated seasonal average is $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.  The variance of this is $\hsshm = A^T \hShy A$.  The standard error of the seasonal average is $\hshm$.  These are all standard results based on approximate multivariate normality of the sampling distribution of $\hbb$.  @Murphy19 gave a similar summary for the purpose of comparing multi-year averages of month-scale spline values for sets of years at the beginning vs. end of a time-series. [I note that equations 2-4 of Murphy et al. appear to be wrong, giving standard least-squares rather than penalized least-squares. I do not necessarily feel we should distract readers with this correction, but we could do so if desired.]

[Here are more detailed notes on @Murphy19.  I think this is more detail than necessary to include in a paper, but I note it here in case it is helpful for clarity among our group.  I'm fairly sure we are both doing the same thing, just with slightly different notation. @Murphy19 used the above relationships to set up a test of the null hypothesis that `y` is the same at two times (each potentialy an average, but not shown here).  To do so, $\bX$ would have two rows, one for each time.  The vector $A^T$ would [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in Chl-A between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.  One can then determine if the difference is significantly different from 0. For me, the most confusing statement of the explanation in @Murphy19 was to call $Z_p$ the "prediction vector".  If I follow their notation, it is the vector of spline basis function evaluations, so it is used to predict `y` but is not itself a set of predictions.]

### Third-stage analysis: Trend analysis of seasonal features with uncertainties

In this section we explain how to use the within-year features ($\hmt$) from the estimated GAMs, and their standard errors $\hshmt$, to investigate across-year trends of within-year features.  In worked examples, we use as features two seasonal averages of Chl-A, one for January-July and the other for August-December.  These periods are chosen for their relevance to California's climate. (In the *Discussion* we consider other features of interest that may be obtained from estimated GAMs.)  For any feature of interest, as long as its estimates come with associated standard errors, we give meta-analysis methods to address two kinds of questions:

1. Do two years, or a group of years, differ significantly in a seasonal average?
2. Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?  This question can be posed in a moving-window manner across a time-series.

[A third question would be "3. Are there significant trends or non-linear fluctuations across years in a seasonal average, where the time-scale of trends is chosen by statistical criteria?"  This needs further work.  My current instinct is that this is too much for this paper, but it should be discussed.]

For all analyses, the response data of interest are $\hmt$, $t = 1, \ldots, N$, with their associated standard errors, $\hshmt$.  $N$ is the number of years of the study.

### Differences among years in seasonal averages

To test the null hypothesis that  $\mu_t = \mu_r$ for years $r$ and $t$, one can use simple normal distribution theory that $\hmt - \hat{\mu}_r$ has standard error $(\hshmt^2 + \hshmr^2)^{1/2}$.  To test the null hypothesis that all means in a group of (>2) years are equal, one can use a simple case of the meta-analysis given in the next section or a general linear hypothesis test. The latter readily includes consideration of multiple testing.  For example, this is provided in R package `multcomp`.  While this package typically uses the results of a linear or generalized linear model fit as input, it is easy to set up the values of $\hmt$ and $\hshmt$ as inputs.  Details are given in the Appendix.

### Linear trends across years at a chosen time-scale of interest

The standard meta-analysis mixed effects model to estimate linear trends is given as follows, using notation similar to [@Sera19].

\begin{equation}
\hmt = \beta_0 + \beta_t t + b_t + \epsilon_t
(\#eq:mixmet)
\end{equation}

where $\beta_0$ is the intercept, $t$ is the year, $\beta_t$ is the slope, $b_t$ is the random effect for year $t$, and $\epsilon_t$ is the residual for year $t$. Here the seasonal average for year $t$ is $\mu_t = \beta_0 + \beta_t t + b_t$.  The "residual", $\epsilon_t$, represents estimation error in $\hmt$, namely $\hmt - \mt$.  This is assumed to be normally disributed with mean 0 and variance $\hsshmt$, which is assumed known from the calculations above.  The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$, what would be considered the "residual" in the sense of unexplained variation not due to the estimation error.  This is assumed normal with mean 0 and unknown variance, $\sbs$, to be estimated.

The R package *mixmeta* was used to estimate this model [@Sera19].  Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.  Following meta-analysis theory, $mixmeta$ handles this model as a linear mixed effects model with some variance components fixed.  The default estimation method is restricted maximum likelihood (REML).

We apply this method for windows of a chosen width of years.  In this way, one can ask whether there is a significant linear trend in Chl-A over a chosen series of years, taking into account all the data in those years.  Some years may have had more sampling, resulting in smaller standard error for $\hmt$, while others may have had less sampling, result in higher standard error. The meta-analysis incorporates such differences in uncertainty magnitudes.

@@

When `select = TRUE` is included, the comparison between models S and SY changes.  This option tells `mgcv` to penalize the linear term's coefficient in the spline.  This would be appropriate if `cont_year` was an explanatory variable subject to variable selection -- which it is not here -- but it would be a strange choice if one is explicitly including a linear term for `cont_year`.  If `select = TRUE` is used, models S and SY would still be effectively equivalent, but it would appear by AIC selection that one model is superior.  That would be an artifact of the strange choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.  A link to an RMarkdown file demonstrating this issue is provided.
